---
title: "RNA-seq Differential Expression Analysis: Ob/Ob (WT vs Het vs Homo)"
author: "govind"
date: "2026-09-16"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

<!--
RECONSTRUCTED 2026-09-23 from the knitted report (Rna_seq.html, rendered 2026-09-16 on macOS,
R 4.4.3 / DESeq2 1.46.0). The original .Rmd was not available. Every R chunk below is the exact
code shown in that report, in order; knitted output, plots and tables are regenerated on render.
Not recoverable from the HTML and therefore re-created: the hidden setup chunk (knitr options and
`tab_dir`, which later chunks write to) and the chunk labels/options (only `install` is known,
with eval = FALSE, from the report text).
Inputs expected next to this file: all_samples_gene_counts.tsv, sample_metadata.csv (not in git;
archived on D:). See ../README.md for known issues affecting interpretation of these results.
-->

```{r setup, include=FALSE}
# Re-created: not shown in the original report.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6)
tab_dir <- "results/tables"
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)
```

# Overview

Differential expression analysis of a mouse RNA-seq experiment across
three genotypes:

- `wt` (n = 2): wt2, wt3
- `het` (n = 2): het2, het3
- `homo` (n = 3): homo1, homo2, homo3

**Primary contrast:** `homo vs wt`.
Secondary: `het vs wt`, `homo vs het`.

**Caveat:** n = 2 for wt and het is the minimum DESeq2
accepts. Dispersion estimates are noisy at this sample size; padj
cutoffs are strict and any hits should be treated as
hypothesis-generating.

# Environment / package check

The intended R environment for this analysis is the conda env
`rnaseq-r` defined in `environment-r.yml` (DESeq2,
edgeR, tximport, apeglm, clusterProfiler, org.Mm.eg.db, pheatmap,
ggplot2, ggrepel, tidyverse, rmarkdown). If you knit outside that env
you’ll hit `there is no package called 'DESeq2'`.

Two ways to fix:

**Option A — activate the conda env, then render
(recommended):**

``` bash
conda env create -f environment-r.yml   # first time only
conda activate rnaseq-r
Rscript -e 'rmarkdown::render("Rna_seq.Rmd")'
```

**Option B — install missing packages into the R you’re
currently using.** Flip `eval = FALSE` to
`eval = TRUE` on the chunk below and knit once; then switch
it back so it doesn’t reinstall every render.

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
cran_pkgs <- c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel")
bioc_pkgs <- c("DESeq2", "apeglm", "clusterProfiler",
               "org.Mm.eg.db", "enrichplot")

missing_cran <- setdiff(cran_pkgs, rownames(installed.packages()))
missing_bioc <- setdiff(bioc_pkgs, rownames(installed.packages()))
if (length(missing_cran)) install.packages(missing_cran)
if (length(missing_bioc)) BiocManager::install(missing_bioc, ask = FALSE,
                                               update = FALSE)
```

## Preflight

```{r}
required <- c("DESeq2", "tidyverse", "pheatmap", "RColorBrewer",
              "ggrepel", "clusterProfiler", "org.Mm.eg.db", "enrichplot")
missing <- required[!vapply(required, requireNamespace,
                            logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing R packages: ", paste(missing, collapse = ", "),
       "\nEither activate the 'rnaseq-r' conda env (see environment-r.yml) ",
       "or set eval=TRUE on the `install` chunk above and re-render.")
}
```

# Libraries

```{r}
suppressPackageStartupMessages({
  library(DESeq2)
  library(pheatmap)
  library(RColorBrewer)
  library(ggrepel)
  library(clusterProfiler)
  library(org.Mm.eg.db)
  library(enrichplot)
  library(tidyverse)   # loaded last so dplyr wins for select/filter/rename
})
# Bulletproof: force dplyr verbs to shadow AnnotationDbi/stats namesakes
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
set.seed(42)
```

# Load data

## Counts matrix

The counts TSV contains gene annotation columns (chr, start, end,
strand, length) followed by the seven sample count columns. We strip
annotation and keep an integer matrix keyed by ENSMUSG gene ID.

```{r}
counts_path <- "all_samples_gene_counts.tsv"
raw <- read.table(counts_path, header = TRUE, sep = "\t",
                  check.names = FALSE, stringsAsFactors = FALSE)

# Save annotation for later (gene length, chr, etc.)
gene_annot <- raw %>% select(gene_id, chr, start, end, strand, length)

count_cols <- c("wt2", "wt3", "het2", "het3", "homo1", "homo2", "homo3")
counts <- raw %>%
  select(all_of(c("gene_id", count_cols))) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()
storage.mode(counts) <- "integer"

cat("Counts matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
```

```{r}
head(counts)
```

## Sample metadata

```{r}
meta <- read.csv("sample_metadata.csv", stringsAsFactors = FALSE) %>%
  # Reorder rows to match count column order
  { .[match(colnames(counts), .$sample), ] } %>%
  mutate(genotype = factor(genotype, levels = c("wt", "het", "homo")))

stopifnot(all(meta$sample == colnames(counts)))
meta
```

# DESeq2 pipeline

## Build DESeqDataSet and pre-filter

Pre-filter genes with fewer than 10 total reads across all samples.
This speeds up the fit and removes obviously uninformative genes without
biasing the test.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ genotype
)

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
cat("Genes retained after filter:", nrow(dds), "\n")
```

## Fit model (joint normalization across all samples)

`DESeq()` computes size factors (median-of-ratios) across
**all** samples jointly, then fits per-gene NB GLMs. This
is the correct way to normalize wt/het/homo together.

```{r}
dds <- DESeq(dds)
sizeFactors(dds)
```

## Normalized counts and VST

```{r}
norm_counts <- counts(dds, normalized = TRUE)
vsd <- vst(dds, blind = FALSE)

write.csv(norm_counts,
          file = file.path(tab_dir, "normalized_counts.csv"))
```

# Global QC

## Library sizes (raw vs normalized)

```{r}
lib_df <- tibble(
  sample   = colnames(counts),
  genotype = meta$genotype,
  raw      = colSums(counts),
  norm     = colSums(norm_counts)
) %>%
  pivot_longer(c(raw, norm), names_to = "type", values_to = "reads")

ggplot(lib_df, aes(sample, reads, fill = genotype)) +
  geom_col() +
  facet_wrap(~ type, scales = "free_y",
             labeller = as_labeller(c(raw = "Raw counts",
                                      norm = "DESeq2-normalized"))) +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Total reads", x = NULL, title = "Library sizes")
```

## Sample-to-sample distance heatmap

```{r}
sample_dist <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dist)
rownames(dist_mat) <- paste(vsd$sample, vsd$genotype, sep = " | ")
colnames(dist_mat) <- vsd$sample

pheatmap(dist_mat,
         clustering_distance_rows = sample_dist,
         clustering_distance_cols = sample_dist,
         col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
         main = "Sample-to-sample distances (VST)")
```

## PCA

```{r}
pca_data <- plotPCA(vsd, intgroup = "genotype", returnData = TRUE)
pct <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = genotype, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  xlab(paste0("PC1: ", pct[1], "% variance")) +
  ylab(paste0("PC2: ", pct[2], "% variance")) +
  theme_bw() +
  ggtitle("PCA (VST-transformed counts)")
```

## Top-500 most variable genes heatmap

```{r}
rv <- rowVars(assay(vsd))
top500 <- order(rv, decreasing = TRUE)[seq_len(min(500, length(rv)))]
mat <- assay(vsd)[top500, ]
mat <- mat - rowMeans(mat)  # center

annot_col <- data.frame(genotype = meta$genotype,
                        row.names = meta$sample)
pheatmap(mat,
         annotation_col = annot_col,
         show_rownames = FALSE,
         scale = "row",
         clustering_distance_cols = "correlation",
         main = "Top 500 most variable genes (VST, row-centered)")
```

## Dispersion estimates

```{r}
plotDispEsts(dds, main = "DESeq2 dispersion estimates")
```

# Differential expression

## Helper: annotate results with gene symbols

```{r}
annotate_res <- function(res_df) {
  ens <- gsub("\\..*$", "", rownames(res_df))
  res_df$ensembl  <- ens
  res_df$symbol   <- mapIds(org.Mm.eg.db, keys = ens, column = "SYMBOL",
                            keytype = "ENSEMBL", multiVals = "first")
  res_df$genename <- mapIds(org.Mm.eg.db, keys = ens, column = "GENENAME",
                            keytype = "ENSEMBL", multiVals = "first")
  res_df$entrez   <- mapIds(org.Mm.eg.db, keys = ens, column = "ENTREZID",
                            keytype = "ENSEMBL", multiVals = "first")
  res_df
}
```

## Helper: MA + Volcano plots

```{r}
plot_volcano <- function(res_df, title, padj_thr = 0.05, lfc_thr = 1,
                         n_label = 20) {
  df <- as.data.frame(res_df) %>%
    filter(!is.na(padj)) %>%
    mutate(sig = case_when(
      padj < padj_thr & log2FoldChange >  lfc_thr ~ "Up",
      padj < padj_thr & log2FoldChange < -lfc_thr ~ "Down",
      TRUE ~ "NS"
    ))
  top <- df %>% filter(sig != "NS") %>%
    arrange(padj) %>% head(n_label)
  ggplot(df, aes(log2FoldChange, -log10(padj), color = sig)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_vline(xintercept = c(-lfc_thr, lfc_thr), lty = 2, color = "grey40") +
    geom_hline(yintercept = -log10(padj_thr),      lty = 2, color = "grey40") +
    scale_color_manual(values = c(Up = "#D7301F", Down = "#2B8CBE", NS = "grey70")) +
    geom_text_repel(data = top, aes(label = symbol),
                    size = 3, max.overlaps = Inf, show.legend = FALSE) +
    theme_bw() +
    labs(title = title, x = "log2 fold change", y = "-log10 adj. p",
         color = NULL)
}
```

## Run all three contrasts

```{r}
run_contrast <- function(dds, num, den, label) {
  res <- results(dds, contrast = c("genotype", num, den), alpha = 0.05)
  # apeglm shrinkage requires coef-based interface; refit with num as reference
  # of the coefficient to shrink, using the built coefficient name.
  coef_name <- paste0("genotype_", num, "_vs_", den)
  if (coef_name %in% resultsNames(dds)) {
    res_shrunk <- lfcShrink(dds, coef = coef_name, type = "apeglm", res = res)
  } else {
    # Fall back to ashr-style normal shrinkage using contrast
    res_shrunk <- lfcShrink(dds, contrast = c("genotype", num, den),
                            type = "normal", res = res)
  }
  df <- as.data.frame(res_shrunk) %>%
    rownames_to_column("gene_id")
  rownames(df) <- df$gene_id
  df <- annotate_res(df)
  df$contrast <- label
  df
}

res_homo_wt   <- run_contrast(dds, "homo", "wt",  "homo_vs_wt")
res_het_wt    <- run_contrast(dds, "het",  "wt",  "het_vs_wt")
res_homo_het  <- run_contrast(dds, "homo", "het", "homo_vs_het")
```

## Summary of DE calls

```{r}
summarize_res <- function(df, padj_thr = 0.05, lfc_thr = 1) {
  df %>%
    filter(!is.na(padj)) %>%
    summarize(
      contrast = unique(contrast),
      tested   = n(),
      sig_strict_up   = sum(padj < padj_thr & log2FoldChange >  lfc_thr),
      sig_strict_down = sum(padj < padj_thr & log2FoldChange < -lfc_thr),
      sig_loose_up    = sum(padj < 0.1 & log2FoldChange >  0),
      sig_loose_down  = sum(padj < 0.1 & log2FoldChange <  0)
    )
}
de_summary <- bind_rows(
  summarize_res(res_homo_wt),
  summarize_res(res_het_wt),
  summarize_res(res_homo_het)
)
de_summary
```

## Write full DE tables

```{r}
write.csv(res_homo_wt,  file.path(tab_dir, "DE_homo_vs_wt.csv"),  row.names = FALSE)
write.csv(res_het_wt,   file.path(tab_dir, "DE_het_vs_wt.csv"),   row.names = FALSE)
write.csv(res_homo_het, file.path(tab_dir, "DE_homo_vs_het.csv"), row.names = FALSE)
```

# Primary contrast: homo vs wt

## MA plot

```{r}
plotMA(results(dds, contrast = c("genotype", "homo", "wt")),
       ylim = c(-6, 6), main = "MA plot: homo vs wt")
```

## Volcano plot

```{r}
plot_volcano(res_homo_wt, "Volcano: homo vs wt")
```

## Top 30 DE genes (by padj)

```{r}
res_homo_wt %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  select(symbol, genename, log2FoldChange, lfcSE, pvalue, padj) %>%
  head(30)
```

## Heatmap of top DE genes across all samples

```{r}
top_genes <- res_homo_wt %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 1) %>%
  arrange(padj) %>%
  head(50) %>%
  pull(gene_id)

if (length(top_genes) >= 2) {
  mat <- assay(vsd)[top_genes, , drop = FALSE]
  rownames(mat) <- ifelse(is.na(res_homo_wt[top_genes, "symbol"]),
                          top_genes, res_homo_wt[top_genes, "symbol"])
  pheatmap(mat,
           annotation_col = annot_col,
           scale = "row",
           show_rownames = TRUE, fontsize_row = 7,
           main = "Top DE genes (homo vs wt), VST z-score")
} else {
  cat("Not enough genes pass strict cutoff for heatmap.\n")
}
```

# Secondary contrast: het vs wt

```{r}
plot_volcano(res_het_wt, "Volcano: het vs wt")
```

```{r}
res_het_wt %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  select(symbol, genename, log2FoldChange, lfcSE, pvalue, padj) %>%
  head(20)
```

# Secondary contrast: homo vs het

```{r}
plot_volcano(res_homo_het, "Volcano: homo vs het")
```

```{r}
res_homo_het %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  select(symbol, genename, log2FoldChange, lfcSE, pvalue, padj) %>%
  head(20)
```

# Cross-contrast comparison

## Overlap of significant DEGs

```{r}
sig <- function(df, padj_thr = 0.05, lfc_thr = 1) {
  df %>% filter(!is.na(padj), padj < padj_thr, abs(log2FoldChange) > lfc_thr) %>%
    pull(gene_id)
}
sig_sets <- list(
  homo_vs_wt  = sig(res_homo_wt),
  het_vs_wt   = sig(res_het_wt),
  homo_vs_het = sig(res_homo_het)
)
sapply(sig_sets, length)
```

```{r}
overlap_df <- tibble(
  set = c("homo_vs_wt only",
          "het_vs_wt only",
          "homo_vs_het only",
          "homo_vs_wt & het_vs_wt",
          "homo_vs_wt & homo_vs_het",
          "het_vs_wt & homo_vs_het",
          "all three"),
  n = c(
    length(setdiff(sig_sets$homo_vs_wt,
                   union(sig_sets$het_vs_wt, sig_sets$homo_vs_het))),
    length(setdiff(sig_sets$het_vs_wt,
                   union(sig_sets$homo_vs_wt, sig_sets$homo_vs_het))),
    length(setdiff(sig_sets$homo_vs_het,
                   union(sig_sets$homo_vs_wt, sig_sets$het_vs_wt))),
    length(setdiff(intersect(sig_sets$homo_vs_wt, sig_sets$het_vs_wt),
                   sig_sets$homo_vs_het)),
    length(setdiff(intersect(sig_sets$homo_vs_wt, sig_sets$homo_vs_het),
                   sig_sets$het_vs_wt)),
    length(setdiff(intersect(sig_sets$het_vs_wt, sig_sets$homo_vs_het),
                   sig_sets$homo_vs_wt)),
    length(Reduce(intersect, sig_sets))
  )
)
ggplot(overlap_df, aes(reorder(set, n), n)) +
  geom_col(fill = "#4C72B0") + coord_flip() + theme_bw() +
  labs(x = NULL, y = "DEGs", title = "DEG overlap across contrasts")
```

## Union heatmap (all DEGs across contrasts)

```{r}
union_genes <- Reduce(union, sig_sets)
if (length(union_genes) >= 2) {
  mat <- assay(vsd)[union_genes, , drop = FALSE]
  labels <- res_homo_wt[union_genes, "symbol"]
  rownames(mat) <- ifelse(is.na(labels) | labels == "",
                          union_genes, labels)
  pheatmap(mat,
           annotation_col = annot_col,
           scale = "row",
           show_rownames = length(union_genes) <= 60,
           fontsize_row = 6,
           main = paste0("Union of DEGs across contrasts (n = ",
                         length(union_genes), ")"))
}
```

# Functional enrichment (GO + KEGG + GSEA)

## Helper

```{r}
run_go <- function(gene_symbols_or_entrez, universe_entrez,
                   ont = "BP", title = "") {
  if (length(gene_symbols_or_entrez) < 10) {
    message("Too few genes for ", title, " (", ont, ")")
    return(NULL)
  }
  enrichGO(gene         = gene_symbols_or_entrez,
           universe     = universe_entrez,
           OrgDb        = org.Mm.eg.db,
           keyType      = "ENTREZID",
           ont          = ont,
           pAdjustMethod = "BH",
           pvalueCutoff = 0.05,
           qvalueCutoff = 0.2,
           readable     = TRUE)
}
```

## Gene universes

```{r}
universe_entrez <- res_homo_wt$entrez %>% na.omit() %>% unique()

de_entrez <- function(df, padj_thr = 0.05, lfc_thr = 1, direction = NULL) {
  x <- df %>% filter(!is.na(padj), padj < padj_thr, abs(log2FoldChange) > lfc_thr)
  if (!is.null(direction)) {
    x <- if (direction == "up")   filter(x, log2FoldChange >  lfc_thr)
         else                     filter(x, log2FoldChange < -lfc_thr)
  }
  x$entrez %>% na.omit() %>% unique()
}
```

## GO enrichment — homo vs wt (primary)

```{r}
up_hw   <- de_entrez(res_homo_wt, direction = "up")
down_hw <- de_entrez(res_homo_wt, direction = "down")
cat("Up genes:", length(up_hw), " Down genes:", length(down_hw), "\n")
```

```{r}
ego_bp_up   <- run_go(up_hw,   universe_entrez, "BP", "up-BP")
ego_bp_down <- run_go(down_hw, universe_entrez, "BP", "down-BP")
ego_mf_up   <- run_go(up_hw,   universe_entrez, "MF", "up-MF")
ego_cc_up   <- run_go(up_hw,   universe_entrez, "CC", "up-CC")

if (!is.null(ego_bp_up) && nrow(ego_bp_up) > 0) {
  print(dotplot(ego_bp_up, showCategory = 15) +
        ggtitle("GO:BP enrichment — upregulated in homo vs wt"))
}
```

```{r}
if (!is.null(ego_bp_down) && nrow(ego_bp_down) > 0) {
  print(dotplot(ego_bp_down, showCategory = 15) +
        ggtitle("GO:BP enrichment — downregulated in homo vs wt"))
}
if (!is.null(ego_mf_up) && nrow(ego_mf_up) > 0) {
  print(barplot(ego_mf_up, showCategory = 15) +
        ggtitle("GO:MF — upregulated in homo vs wt"))
}
```

```{r}
if (!is.null(ego_cc_up) && nrow(ego_cc_up) > 0) {
  print(barplot(ego_cc_up, showCategory = 15) +
        ggtitle("GO:CC — upregulated in homo vs wt"))
}
```

## KEGG pathway enrichment — homo vs wt

`enrichKEGG` calls the KEGG REST API and can hang
indefinitely if the server is slow/offline. We wrap each call in a child
process with a hard 120-second timeout so a KEGG outage can’t block the
render.

```{r}
options(timeout = 60)  # affects base R HTTP timeouts

run_kegg <- function(genes, universe, label, timeout_s = 120) {
  if (length(genes) < 10) {
    message("Too few genes for KEGG (", label, ")"); return(NULL)
  }
  # setTimeLimit throws an interrupt at next R-level check if elapsed exceeds
  # timeout_s. Combined with `options(timeout)` for HTTP, this bounds KEGG runtime.
  tryCatch({
    setTimeLimit(elapsed = timeout_s, transient = TRUE)
    on.exit(setTimeLimit(elapsed = Inf, transient = TRUE), add = TRUE)
    enrichKEGG(gene = genes, organism = "mmu",
               universe = universe, pvalueCutoff = 0.05)
  }, error = function(e) {
    message("KEGG failed (", label, "): ", e$message); NULL
  })
}

kegg_up   <- run_kegg(up_hw,   universe_entrez, "up-homo-vs-wt")
kegg_down <- run_kegg(down_hw, universe_entrez, "down-homo-vs-wt")

# Some enrichment steps consume tempfiles heavily; ensure tempdir is intact
# before subsequent chunks so `file("", "w+b")` can't fail.
if (!dir.exists(tempdir())) dir.create(tempdir(), recursive = TRUE)

if (!is.null(kegg_up) && nrow(kegg_up) > 0) {
  print(dotplot(kegg_up, showCategory = 15) +
        ggtitle("KEGG — up in homo vs wt"))
}
```

```{r}
if (!is.null(kegg_down) && nrow(kegg_down) > 0) {
  print(dotplot(kegg_down, showCategory = 15) +
        ggtitle("KEGG — down in homo vs wt"))
}
```

## GSEA (ranked by shrunk log2FC) — homo vs wt

```{r}
if (!dir.exists(tempdir())) dir.create(tempdir(), recursive = TRUE)

rank_vec <- res_homo_wt %>%
  filter(!is.na(entrez), !is.na(log2FoldChange)) %>%
  distinct(entrez, .keep_all = TRUE) %>%
  { setNames(.$log2FoldChange, .$entrez) } %>%
  sort(decreasing = TRUE)

gse_bp <- tryCatch(
  withCallingHandlers(
    gseGO(geneList = rank_vec, OrgDb = org.Mm.eg.db, keyType = "ENTREZID",
          ont = "BP", pvalueCutoff = 0.05, verbose = FALSE),
    warning = function(w) invokeRestart("muffleWarning")),
  error = function(e) { message("GSEA failed: ", e$message); NULL }
)

if (!is.null(gse_bp) && nrow(gse_bp) > 0) {
  gse_bp <- setReadable(gse_bp, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  print(dotplot(gse_bp, showCategory = 15, split = ".sign") +
        facet_grid(. ~ .sign) +
        ggtitle("GSEA GO:BP — homo vs wt"))
}
```

## Concept network (top GO:BP terms)

```{r}
if (!is.null(ego_bp_up) && nrow(ego_bp_up) >= 3) {
  fc_named <- setNames(res_homo_wt$log2FoldChange, res_homo_wt$entrez)
  fc_named <- fc_named[!is.na(names(fc_named))]
  tryCatch({
    print(cnetplot(ego_bp_up, showCategory = 5, color.params = list(foldChange = fc_named)) +
          ggtitle("cnetplot — GO:BP upregulated (homo vs wt)"))
  }, error = function(e) {
    message("cnetplot skipped: ", e$message)
    print(cnetplot(ego_bp_up, showCategory = 5) +
          ggtitle("cnetplot — GO:BP upregulated (homo vs wt)"))
  })
  tryCatch({
    print(emapplot(pairwise_termsim(ego_bp_up), showCategory = 20) +
          ggtitle("emapplot — GO:BP upregulated (homo vs wt)"))
  }, error = function(e) message("emapplot skipped: ", e$message))
}
```

## GO enrichment — het vs wt (secondary, condensed)

```{r}
up_ew   <- de_entrez(res_het_wt, direction = "up")
down_ew <- de_entrez(res_het_wt, direction = "down")

ego_bp_up_ew   <- run_go(up_ew,   universe_entrez, "BP", "het-up-BP")
ego_bp_down_ew <- run_go(down_ew, universe_entrez, "BP", "het-down-BP")

if (!is.null(ego_bp_up_ew) && nrow(ego_bp_up_ew) > 0) {
  print(dotplot(ego_bp_up_ew, showCategory = 15) +
        ggtitle("GO:BP — up in het vs wt"))
}
```

```{r}
if (!is.null(ego_bp_down_ew) && nrow(ego_bp_down_ew) > 0) {
  print(dotplot(ego_bp_down_ew, showCategory = 15) +
        ggtitle("GO:BP — down in het vs wt"))
}
```

## Save enrichment tables

```{r}
save_enrich <- function(x, name) {
  if (!is.null(x) && nrow(x) > 0) {
    write.csv(as.data.frame(x),
              file.path(tab_dir, paste0("enrich_", name, ".csv")),
              row.names = FALSE)
  }
}
save_enrich(ego_bp_up,     "GO_BP_up_homo_vs_wt")
save_enrich(ego_bp_down,   "GO_BP_down_homo_vs_wt")
save_enrich(ego_mf_up,     "GO_MF_up_homo_vs_wt")
save_enrich(ego_cc_up,     "GO_CC_up_homo_vs_wt")
save_enrich(kegg_up,       "KEGG_up_homo_vs_wt")
save_enrich(kegg_down,     "KEGG_down_homo_vs_wt")
save_enrich(gse_bp,        "GSEA_BP_homo_vs_wt")
save_enrich(ego_bp_up_ew,  "GO_BP_up_het_vs_wt")
save_enrich(ego_bp_down_ew,"GO_BP_down_het_vs_wt")
```

# Cross-check with imaged markers

These genes were previously imaged in the same oocyte cohort. This
section extracts their normalized counts and DE stats across all three
contrasts so you can cross-verify the imaging phenotype against the
transcript level.

Notes:

- **H3K9me2** is a histone mark, not a gene. We report
  the main writers (`Ehmt2` = G9a, `Ehmt1` = GLP)
  and erasers (`Kdm3a/b`, `Kdm4a/b/c`).
- **L1ORF1** — LINE-1 ORF1 is a retrotransposon sequence,
  not a standard annotated gene. Standard gene-level RNA-seq (like this
  pipeline) does not quantify repeat elements. `L1td1` (LINE-1
  type transposase domain containing 1) is included as the closest
  annotated proxy; for true LINE-1 activity you’d need a TE-aware
  re-quantification (e.g., TEtranscripts).

```{r}
imaged_markers <- tibble::tribble(
  ~imaged,     ~symbol,   ~role,
  "FTO",       "Fto",     "m6A demethylase (imaged)",
  "gH2AX",     "H2ax",    "DNA-damage marker (imaged)",
  "Rad51",     "Rad51",   "HR repair (imaged)",
  "HP1g",      "Cbx3",    "Heterochromatin protein 1 gamma (imaged)",
  "H3K9me2",   "Ehmt2",   "H3K9 methyltransferase - G9a (writer)",
  "H3K9me2",   "Ehmt1",   "H3K9 methyltransferase - GLP (writer)",
  "H3K9me2",   "Kdm3a",   "H3K9me1/2 demethylase (eraser)",
  "H3K9me2",   "Kdm3b",   "H3K9me1/2 demethylase (eraser)",
  "H3K9me2",   "Kdm4a",   "H3K9me2/3 demethylase (eraser)",
  "H3K9me2",   "Kdm4b",   "H3K9me2/3 demethylase (eraser)",
  "H3K9me2",   "Kdm4c",   "H3K9me2/3 demethylase (eraser)",
  "L1ORF1",    "L1td1",   "LINE-1 type transposase domain (proxy)"
)
imaged_markers
```

## Combined DE stats across contrasts

```{r}
# Pull each gene from each contrast, then pivot wide by contrast.
pick <- function(res_df, contrast_lbl) {
  res_df %>%
    filter(symbol %in% imaged_markers$symbol) %>%
    select(symbol, log2FoldChange, padj, baseMean) %>%
    rename(!!paste0("lfc_", contrast_lbl)  := log2FoldChange,
           !!paste0("padj_", contrast_lbl) := padj)
}

de_hw <- pick(res_homo_wt,  "homo_vs_wt")  %>% select(-baseMean)
de_ew <- pick(res_het_wt,   "het_vs_wt")   %>% select(-baseMean)
de_hh <- pick(res_homo_het, "homo_vs_het")
# baseMean is the same across contrasts (single fit) so keep it from one

imaged_de <- imaged_markers %>%
  left_join(de_hh %>% select(symbol, baseMean), by = "symbol") %>%
  left_join(de_hw, by = "symbol") %>%
  left_join(de_ew, by = "symbol") %>%
  left_join(de_hh %>% select(-baseMean), by = "symbol") %>%
  mutate(across(starts_with("lfc_"),  ~ round(.x, 3)),
         across(starts_with("padj_"), ~ signif(.x, 3)),
         baseMean = round(baseMean, 1))
imaged_de
```

```{r}
write.csv(imaged_de, file.path(tab_dir, "imaged_markers_DE.csv"),
          row.names = FALSE)
```

## Per-sample normalized counts

```{r}
sym_to_id <- res_homo_wt %>%
  filter(symbol %in% imaged_markers$symbol) %>%
  select(gene_id, symbol) %>%
  distinct(symbol, .keep_all = TRUE)

nc <- norm_counts[sym_to_id$gene_id, , drop = FALSE]
rownames(nc) <- sym_to_id$symbol

imaged_counts <- as.data.frame(nc) %>%
  tibble::rownames_to_column("symbol") %>%
  left_join(imaged_markers %>% select(symbol, imaged, role), by = "symbol") %>%
  select(imaged, symbol, role, all_of(colnames(counts))) %>%
  mutate(across(all_of(colnames(counts)), ~ round(.x, 1)))
imaged_counts
```

```{r}
write.csv(imaged_counts,
          file.path(tab_dir, "imaged_markers_normalized_counts.csv"),
          row.names = FALSE)
```

## Heatmap of imaged markers (VST, row-centered)

```{r}
have <- sym_to_id$gene_id
if (length(have) >= 2) {
  mat <- assay(vsd)[have, , drop = FALSE]
  rownames(mat) <- sym_to_id$symbol
  pheatmap(mat,
           annotation_col = annot_col,
           scale = "row",
           cluster_rows = length(have) > 2,
           main = "Imaged markers — VST z-score across samples")
} else {
  cat("Not enough imaged markers found in the dataset.\n")
}
```

## Genes not found in the dataset

```{r}
missing_syms <- setdiff(imaged_markers$symbol, sym_to_id$symbol)
if (length(missing_syms) == 0) {
  cat("All imaged-marker gene symbols were found in the DE tables.\n")
} else {
  cat("Not found (may be filtered out by low-count filter):\n")
  print(missing_syms)
}
```

# Session info

```{r}
sessionInfo()
```
